\documentclass{beamer}
\usepackage{amsmath, minted, tikz}
\usepackage{mathtools, textcomp, qtree, graphicx}
\usetheme{Berkeley}
\usecolortheme{mbhs}

\title{Welcome to Computer Team!}
\subtitle{Introduction to Computer Systems}

\author{Noah Singer}
\institute{Montgomery Blair High School Computer Team}

\begin{document}

\begin{frame}
\titlepage
\end{frame}

\section{Introduction}

\begin{frame}{Hey there!}
Welcome to the Montgomery Blair High School Computer Team for the 2017-18 school year!
\vspace{.1in}
\\
We're excited to discuss and teach all sorts of topics and algorithms this year.
\vspace{.1in}
\\
Computer science is an awesome subject, and it's fun to study both as a career, hobby, or casual interest.
\vspace{.1in}
\\
Let's dive right in!
\end{frame}

\begin{frame}{Computer Team}
\textbf{What computer team is...}
\begin{itemize}
\item A place to hang out and eat lunch every Thursday
\item A cool discussion group for computer science
\item An award-winning competition programming group
\item A place for people to learn from each other
\item Open to people of all experience levels
\end{itemize}
\vspace{.2in}
\textbf{... what computer team isn't}
\begin{itemize}
\item A programming club
\item A highly competitive environment
\item A rigorous course that requires commitment
\end{itemize}
\end{frame}

\begin{frame}{Definitions}
\begin{enumerate}
\item What is a computer?
\pause
\begin{definition}[Computer]
A programmable. usually electronic device that can store, retrieve, and process data.
\end{definition}
\vspace{0.2in}
\pause
\item What is an algorithm?
\pause
\begin{definition}[Algorithm]
A step-by-step procedure for solving a problem or accomplishing some end, especially by a computer.
\end{definition}
\end{enumerate}
\end{frame}

\section{Topics}

\begin{frame}{Algorithms}
Fundamental question: \textit{What is the best way to solve a given problem?}
\vspace{0.1in}

\begin{itemize}
\item Graph algorithms
\item Dynamic programming
\item Computational geometry
\item Greedy algorithms
\item Divide-and-conquer and recursive algorithms
\item Online algorithms
\end{itemize}
\end{frame}

\begin{frame}{Theory of computation}
Fundamental question: \textit{What can a computer do?}
\vspace{0.1in}

\begin{itemize}
\item Computational complexity theory
\item Formal language theory
\item Automata
\item Computability theory
\item Models of computation
\begin{itemize}
\item Lambda calculus
\item Turing machines
\end{itemize}
\item Reductions
\end{itemize}
\end{frame}

\begin{frame}{Machine learning+}
Fundamental question: \textit{How can we get computers to solve complex problems?}
\vspace{0.1in}

\begin{itemize}
\item Linear and logistic regression
\item Neural networks
\item Support vector machines
\item Natural language processing
\item Computer vision
\item Speech recognition and synthesis
\item Fourier analysis and signal processing
\item Convex optimization (my summer research!)
\end{itemize}
\end{frame}

\begin{frame}{Systems}
Fundamental question: \textit{How do computers function at a low level?}
\vspace{0.1in}

\begin{itemize}
\item Compilers
\item Computer architecture
\item Operating system design
\item Memory managers and task schedulers
\end{itemize}
\end{frame}

\begin{frame}{Contest preparation}
Fundamental question: \textit{How can we beat all our rival high schools (TJ) at programming?}
\vspace{0.1in}

\begin{itemize}
\item PClassic
\item USACO
\item UVA
\item VCU
\item UMD
\end{itemize}
\end{frame}




\end{document}